Computable Compressed Matrices\\Suplementary Material

### 2013-10-02 suppl.tex
Suplemento revisado...

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{cite}

@@

In this supplementary material we will explore basic arithmetic operations
(addition, subtraction, division and multiplication) over the elements of
bitstring compressed arrays. We will consider integer matrices compressed via
both the SM and the VLB methods.

In these operations, the most important aspect is the bit-length of the result
in relation to those of the operands. As the operations are all done in
binary, when the result bit-length increases, the resulting matrix will require
more space to store and the operation in itself gets more complex, as in place
operations are not possible. In the following examples we will explore these
kinds of operations.

@@

In base 2 the carry over  behavior is the same as we observer when operating
with decimals, only the base is different. Adding $1+1$ results in $0$ and a
carry over of 1, this is similar to the decimal sum $5+5$ where we also have a
carry over of 1. Other single digit additions in base 2 are: $1 + 0
= 1$, $0 + 1 = 1$, and $0 + 0 = 0$. Consider the following addition $n_1+n_2$
where $n_1=7$ and $n_2=10$. This addition is illustrated in table \ref{tab:01}
in both decimal and binary. Note that when we add $1+1$, a carry over is
generated to be added to bit on the left. In total we generate three bits in
carry over.

In Table \ref{tab:02} we show the addition of $7 + 7$ which generates a 4 bits
long number. In integer additions, the results may, in the very least, be as
long (in bits) as the greatest operand (for example $1+2=3$), but is frequently
longer.

\begin{table}[ht]
	\centering
    \caption{Comparing binary and decimal sum ($7+10$).}
    \begin{tabular}{crrrc}
    \hline
    	  & Decimal & & Binary & bit-length \\
    \hline      
    Carry Over& 	& & \textit{11100} & \\
    $n_1$ & 7   & & 111 & 3 \\
    $n_2$ & \underline{10}	& & \underline{1010} & 4 \\
    Total & 17	& & 10001 & 5 \\
    \hline
	\end{tabular}
    \label{tab:01}
\end{table}

\begin{table}[ht]
	\centering
    \caption{Comparacao da soma em base decimal e binaria.}
    \begin{tabular}{crrrc}
    \hline
    	  & Decimal & & Binario & Bit-length \\
    \hline      
    Carry & 	& & \textit{1110} & \\
    $n_1$ & 7   & & 111 & 3 \\
    $n_1$ & \underline{~7}	& & \underline{111} & 3 \\
    Total & 14	& & 1110 & 4 \\
    \hline
	\end{tabular}
    \label{tab:02}
\end{table}

\paragraph{Subtraction}
Depending of the number involved in this operation, the result may be negative.
This fact alone generates the need for an extra bit to represent the sign ($+$
or
$-$). The usual way to achieve this is to use the  two's complement
\cite{flores1963logic} representation which has the advantage of making the
addition subtraction and multiplication operation the same as those for unsigned
binary numbers. This algorithm is what is actually implemented in CPUs. To
exemplify let's subtract $7$ from $10$. To make use of the
same algorithm of the sum , we represent the operation as $10+(-7)$. First we
need to convert the operands to two's complement representation (C2):

@@

In multiplication the behavior of the bit-length is different from the sum and
subtraction. In the worst case the product has a bit-length which is the sum of
the bit-lengths of the factors. For example, in the product $7\times7$, the
product is 6 bits long: $49$ which in binary is $110001$.

@@

So far we have examined the 4 fundamental arithmetic operations. In summary the
implication for memory allocation of the results are the folowing, in the worst
case scenarios:

\begin{itemize}
\item Addition: requires 1 extra bit above the bit-length of the greatest
operand;
\item Subtraction: Requires the same number of bits as the greatest
operand;
\item Multiplication: Requires the sum of the bit-lenghts of the
operands;
\item Division: Requires the same bit-length as the dividend;
\end{itemize}

Figure \ref{fig:bitlength} shows the bit-lengths of results of the four
operations with integer operands up to 8 bits in length. The conclusions listed
above, are visually emphasized in the figures.

@@

\begin{figure}[h]
 \centering
%  \includegraphics[width=12cm]{./bitlength.png}
 % simples.png: 4800x3600 pixel, 600dpi, 20.32x15.24 cm, bb=0 0 576 432
 \caption{Bit-length of the results for each of the four operations. The axes
represent the values of the operands.}
 \label{fig:bitlength}
\end{figure}

@@

e

@@

Consider now just the bitstring stored in the first line of each matrix ($AA$
and $BB$). To recover the original elements we need to now the bit-lengthof
the blocks of memory containing them, for the SM method they are all the same
length. To obtain the values in an efficient way, we will use a binary mask.
The mask will recover the first and third blocks at the same time to save time.
The mask is stored as a bitstring the same length as those storing the matrix
elements. The mask for the first and third elements of matrix $AA$ is depicted
in \ref{eq:06}. To recover the second and fouth elements, we apply the mask
shown on \ref{eq:07}. For Matrix $BB$, see \ref{eq:08} and \ref{eq:09} for the
respective masks.

@@

Once defined the mask we can apply it to matrices $A$ and $B$. We use the mask
by applying the boolean function AND. Let tempA and tempB the recovered
elements. The recovering process is illustrated below. Note that only the
positions where the mask is $1$ are retained.

%\begin{equation}\label{eq:10}
  \begin{align*}
   A[1,1]			&=	\ldots0001|0011|0101|1000\\
   \text{mask A}	&=	\ldots0000|1111|0000|1111\\
   tempA 			&=	\ldots0000|0011|0000|1000
  \end{align*}
%\end{equation}

%\begin{equation}\label{eq:11}
  \begin{align*}
   B[1,1]			&= \ldots00001|00011|00111|01001 \\
   \text{mask B}	&= \ldots00000|11111|00000|11111 \\
   tempB 			&= \ldots00000|00011|00000|01001
  \end{align*}
%\end{equation}

@@

Suppose we want to add both matrices and store it in a matrix $S$. for that the
blocks of tempA and tempB must have the same length. As for the addition we
need, in the worst case, an extra bit, the block length of the resulting
bitstring will be 6. The string from each matrix, padded on the left towards
this new length,  are shown in equations \ref{eq:12} and \ref{eq:13}.

@@

Therefore, as the final result of the operation $S= A +
B$ we have:

\begin{align}
	S &= \begin{bmatrix}
			549649\\ 
			274889\\ 
			6181644
		\end{bmatrix}
        =
         \begin{bmatrix}
			\ldots000010000110001100010001\\
			\ldots000001000011000111001001\\
			\ldots010111100101001100001100
		\end{bmatrix}
      \\ \nonumber
      &= \begin{bmatrix}
\ldots\underbrace{000010}_{2}\underbrace{000110}_{6}\underbrace{001100}_{12}
\underbrace{010001}_{17}\\	
 \ldots\underbrace{001101}_{13}\underbrace{010001}_{17}\underbrace{010100}_{20}
\underbrace{011000}_{24}\\
 \ldots\underbrace{010111}_{23}\underbrace{100101}_{37}\underbrace{001100}_{12}
\underbrace{001100}_{12}
		\end{bmatrix}
\end{align}

\paragraph{Subtraction}
To perform the subtraction, we need to use C2 notation. Supose  we need to
calculate $D = B - A$. This operation can be done in the same way, by
converting it to an addition: $D = B + (-A)$, where $-A$ will be converted to
C2 (equation \ref{eq:ac2}). Note the elements will be calculated with six
digits.

\begin{align}\label{eq:ac2}
	-A &= \begin{bmatrix}
			4952\\ 
			52841\\ 
			14251
		\end{bmatrix} 
        =
        \begin{bmatrix}
   			\ldots111111111101111011111000\\ 
			\ldots110100110010111010110111\\ 
			\ldots111101111001110110110101
		\end{bmatrix}
        \\\nonumber
        &= \begin{bmatrix}
 \ldots\underbrace{111111}_{-1}\underbrace{111101}_{-3}\underbrace{111011}_{-5}
\underbrace{111000}_{-8}\\
\ldots\underbrace{110100}_{-12}\underbrace{110010}_{-14}\underbrace{111010}_{-6}
\underbrace{110111}_{-9}\\
\ldots\underbrace{111101}_{-3}\underbrace{111001}_{-7}\underbrace{110110}_{-10}
\underbrace{110101}_{-11}
		\end{bmatrix}
\end{align}

@@

To exemplify, let's examine closely the operations $B[1,1] + (-A[1,1])$,
$B[2,1] + (-A[2,1])$ and $B[3,1] + (-A[3,1])$. We use tempB to store each line
of the matrix B converted to C2. The numbers $1$ marked in the bitstring, are
overflows and must be removed. The result is shown below:

@@

\begin{align*}
tempB	  &= \ldots010100|011110|000010|000001 \\
+(-A[3,1])&= \ldots111101|111001|110110|110101\\ 
D[3,1]	  &= \ldots\sout{1}010001|\sout{1}110111|\sout{1}111000|110110
\end{align*}

@@

\begin{align}
	D &= \begin{bmatrix}
			129\\ 
			14111238\\ 
			4685366
		\end{bmatrix} 
        =
        \begin{bmatrix}
   			\ldots000000000000000010000001 \\
			\ldots110101110101001000000110 \\
			\ldots010001110111111000110110
		\end{bmatrix}
      \\\nonumber
        &=  \begin{bmatrix}
 \ldots\underbrace{000000}_{0}\underbrace{000000}_{0}\underbrace{000010}_{2}
\underbrace{000001}_{1} \\ 			 
 \ldots\underbrace{110101}_{-11}\underbrace{110101}_{-11}\underbrace{001000}_{8}
\underbrace{000110}_{6} \\
 \ldots\underbrace{010001}_{17}\underbrace{110111}_{23}\underbrace{111000}_{-8}
\underbrace{110110}_{-10}
		\end{bmatrix}
\end{align}

\paragraph{Multiplication}
Let's now examine matrix multiplication. Consider the product $P = A \times
B^{t}$, where $B^t$ is the transpose of B. Thus the product becomes:

@@

\begin{equation}\label{eq:prodbs}
	P = AA \times BB = \begin{bmatrix}
			4952\\ 
			52841\\ 
			14251
		\end{bmatrix} 
        \times\times
        \begin{bmatrix}
			36073 & 36303 & 686145
		\end{bmatrix}
\end{equation}

\begin{equation}
	P = AA \times BB = 
    	\begin{bmatrix}

\ldots\underbrace{0001}_{1}\underbrace{0011}_{3}\underbrace{0101}_{5}\underbrace
{1000}_{8}\\

\ldots\underbrace{1100}_{12}\underbrace{1110}_{14}\underbrace{0110}_{6}
\underbrace{1001}_{9}\\

\ldots\underbrace{0011}_{3}\underbrace{0111}_{7}\underbrace{1010}_{10}
\underbrace{1011}_{11}
		\end{bmatrix}
\end{equation}
\begin{small}
\begin{equation}
        \times\times \\
        \begin{bmatrix}\mathsf{
\ldots\underbrace{00001}_{1}\underbrace{00011}_{3}\underbrace{00111}_{7}
\underbrace{01001}_{9}}  &\mathsf{
\ldots\underbrace{00001}_{1}\underbrace{00011}_{3}\underbrace{01110}_{14}
\underbrace{01111}_{15}} & \mathsf{
\ldots\underbrace{10100}_{20}\underbrace{11110}_{30}\underbrace{00010}_{2}
\underbrace{00001}_{1}}
        \end{bmatrix}\nonumber
\end{equation}
\end{small}
To calculate $P[1,1]$ which in uncompressed form is $1 \times 1 + 3 \times 3 + 5
\times 7 + 8 \times 9$, we first must extract the numbers from the bitstrings
and then proceed with the linear combination. The extraction will make use of
masks as described before. Let tempA and tempB store the values of blocks 1 and
3 of AA[1,1] and BB[1,1]. Now we need to determine the size of the blocks
(bit-length) containing each element of the resulting matrix. As demonstrated
before, in the worst case, the product will have a bit-length which is the sum
of the bit-lengths of the operands. For this example this length is $4+5=9$,
however, we also have three additions for each element,adding a total of 3
extra bits, thus we end up with a bit-length of 12 for each element of the
product. Having determined the bit-length of the product, we can now we can do
the actual calculations and store the results.

@@

Again we can see that P is already in compressed form, which confirms that the
entire operation was conducted without decompressing the data.

>>>>

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{todonotes}

@@

In this supplementary material we will explore basic arithmetic operations
(addition, subtraction, division and multiplication) over the elements of
bitstring compressed arrays. We will consider integer matrices compressed via
both the SM and the VLB methods, \todo{inseri esse texto} but the proposed 
methods can be applied to integers and real numbers with sign.

In these operations, the most important aspect is the bit-length of the result
in relation to those of the operands. As the operations are all done in
binary, when the result bit-length increases, the resulting matrix will require
more space to store and the operation in itself gets more complex, as in place
operations are not possible. In the following examples, we will explore these
kinds of operations.

@@

In base 2 the carry over behavior is the same as we observer when operating
with decimals, only the base is different. Adding $1+1$ results in $0$ and a
carry over of 1, this is similar to the decimal sum $5+5$ where we also have a
carry over of 1. Other single digit additions in base 2 are: $1 + 0
= 1$, $0 + 1 = 1$, and $0 + 0 = 0$. Consider the following addition $n_1+n_2$
where $n_1=7$ and $n_2=10$. This addition is illustrated in table \ref{tab:01}
in both decimal and binary. Note that, when we add $1+1$, a carry over is
generated to be added to bit on the left. In total, we generate three bits in
carry over.

In Table \ref{tab:02}, we show the addition of $7 + 7$ which generates a 4 bits
long number. In integer additions, the results may, in the very least, be as
long (in bits) as the greatest operand (for example $1+2=3$), but is frequently
longer.

\begin{table}[ht]
	\centering
    \caption{Comparing binary and decimal sum ($7+10$).}
    \begin{tabular}{crrrc}
    \hline
    	  & Decimal & & Binary & bit-length \\
    \hline      
    Carry Over& 	& & \textit{11100} & \\
    $n_1$ & 7   & & 111 & 3 \\
    $n_2$ & \underline{10}	& & \underline{1010} & 4 \\
    Total & 17	& & 10001 & 5 \\
    \hline
	\end{tabular}
    \label{tab:01}
\end{table}
\todo{Traduzi o título que estava em português.}
\begin{table}[ht]
    \centering
    \caption{Comparison of the sum in binary and decimal base.}
    \begin{tabular}{crrrc}
    \hline
    	  & Decimal & & Binario & Bit-length \\
    \hline      
    Carry & 	& & \textit{1110} & \\
    $n_1$ & 7   & & 111 & 3 \\
    $n_1$ & \underline{~7}	& & \underline{111} & 3 \\
    Total & 14	& & 1110 & 4 \\
    \hline
	\end{tabular}
    \label{tab:02}
\end{table}

\paragraph{Subtraction}

Depending of the number involved in this operation, the result may be negative.
This fact alone generates the need for an extra bit to represent the sign ($+$
or $-$). The usual way to achieve this is to use the  two's complement
\cite{flores1963logic} representation which has the advantage of making the
addition subtraction and multiplication operation the same as those for unsigned
binary numbers. This algorithm \todo{removi is what} is actually implemented in 
CPUs. To exemplify let's subtract $7$ from $10$. To make use of the
same algorithm of the sum , we represent the operation as $10+(-7)$. First we
need to convert the operands to two's complement representation (C2):

@@

In multiplication the behavior of the bit-length is different from the sum and
subtraction. In the worst case, the product has a bit-length which is the sum of
the bit-lengths of the factors. For example, in the product $7\times7$, the
product is 6 bits long: $49$ which in binary is $110001$.

@@

So far we have examined the 4 fundamental arithmetic operations. In summary the
implication for memory allocation of the results are the following, in the worst
case scenarios:

\begin{itemize}
\item Addition: requires 1 extra bit above the bit-length of the greatest
operand;
\item Subtraction: Requires the same number of bits as the greatest
operand;
\item Multiplication: Requires the sum of the bit-lengths of the
operands;
\item Division: Requires the same bit-length as the dividend;
\end{itemize}

\todo{adicionei esse trecho - As can be seen in} As can be seen in figure 
\ref{fig:bitlength}, the bit-lengths of results of the four
operations with integer operands up to 8 bits in length. The conclusions listed
above, are visually emphasized in the figures.

@@

\begin{figure}[h]
 \centering
 \includegraphics[width=12cm]{./bitlength.png}
 \caption{Bit-length of the results for each of the four operations. The axes
represent the values of the operands.}
 \label{fig:bitlength}
\end{figure}

@@

and

@@

Consider now just the bitstring stored in the first line of each matrix ($AA$
and $BB$). To recover the original elements we need to now the bit-length of
the blocks of memory containing them, for the SM method they are all the same
length. To obtain the values in an efficient way, we will use a binary mask.
The mask will recover the first and third blocks at the same time to save time.
The mask is stored as a bitstring the same length as those storing the matrix
elements. The mask for the first and third elements of matrix $AA$ is depicted
in \ref{eq:06}. To recover the second and fourth elements, we apply the mask
shown on \ref{eq:07}. For Matrix $BB$, see \ref{eq:08} and \ref{eq:09} for the
respective masks.

@@

Once defined the mask we can apply it to matrices $A$ and $B$. We use the mask
by applying the boolean function AND. Let $tempA$ and $tempB$ the recovered
elements. The recovering process is illustrated below. Note that only the
positions where the mask is $1$ are retained.

  \begin{align*}
   A[1,1]		&=	\ldots0001|0011|0101|1000\\
   \text{mask A}	&=	\ldots0000|1111|0000|1111\\
   tempA 		&=	\ldots0000|0011|0000|1000
  \end{align*}

  \begin{align*}
   B[1,1]		&= \ldots00001|00011|00111|01001 \\
   \text{mask B}	&= \ldots00000|11111|00000|11111 \\
   tempB 		&= \ldots00000|00011|00000|01001
  \end{align*}

@@

Suppose we want to add both matrices and store it in a matrix $S$. for that the
blocks of $tempA$ and $tempB$ must have the same length. As for the addition we
need, in the worst case, an extra bit, the block length of the resulting
bitstring will be 6. The string from each matrix, padded on the left towards
this new length,  are shown in equations \ref{eq:12} and \ref{eq:13}.

@@

Therefore, as the final result of the operation $S= A + B$, we have:

\begin{align}
	S &=\begin{bmatrix}
	      549649\\ 
	      274889\\ 
	      6181644
	    \end{bmatrix}
        =
        \begin{bmatrix}
	    \ldots000010000110001100010001\\
	    \ldots000001000011000111001001\\
	    \ldots010111100101001100001100
	\end{bmatrix}
      \\ \nonumber      
      &=\begin{bmatrix}
  \ldots\underbrace{000010}_{2}\underbrace{000110}_{6}\underbrace{001100}_{12}
  \underbrace{010001}_{17}\\	
  \ldots\underbrace{001101}_{13}\underbrace{010001}_{17}\underbrace{010100}_{20}
  \underbrace{011000}_{24}\\
  \ldots\underbrace{010111}_{23}\underbrace{100101}_{37}\underbrace{001100}_{12}
  \underbrace{001100}_{12}
	\end{bmatrix}
\end{align}

\paragraph{Subtraction}
To perform the subtraction, we need to use C2 notation. Suppose  we need to
calculate $D = B - A$. This operation can be done in the same way, by
converting it to an addition: $D = B + (-A)$, where $-A$ will be converted to
C2 (equation \ref{eq:ac2}). Note the elements will be calculated with six
digits.

\begin{align}\label{eq:ac2}
	-A &= \begin{bmatrix}
		4952\\ 
		52841\\ 
		14251
	      \end{bmatrix} 
        =
        \begin{bmatrix}
	  \ldots111111111101111011111000\\ 
	  \ldots110100110010111010110111\\ 
	  \ldots111101111001110110110101
	\end{bmatrix}
        \\\nonumber
        &=\begin{bmatrix}
\ldots\underbrace{111111}_{-1}\underbrace{111101}_{-3}\underbrace{111011}_{-5}
\underbrace{111000}_{-8}\\
\ldots\underbrace{110100}_{-12}\underbrace{110010}_{-14}\underbrace{111010}_{-6}
\underbrace{110111}_{-9}\\
\ldots\underbrace{111101}_{-3}\underbrace{111001}_{-7}\underbrace{110110}_{-10}
\underbrace{110101}_{-11}
	  \end{bmatrix}
\end{align}

@@

To exemplify, let's examine closely the operations $B[1,1] + (-A[1,1])$,
$B[2,1] + (-A[2,1])$ and $B[3,1] + (-A[3,1])$. We use $tempB$ to store each line
of the matrix B converted to C2. The numbers $1$ marked in the bitstring, are
overflows and must be removed. The result is shown below:

@@

\begin{align*}
    tempB	&= \ldots010100|011110|000010|000001 \\
    +(-A[3,1])	&= \ldots111101|111001|110110|110101\\ 
    D[3,1]	&= \ldots\sout{1}010001|\sout{1}110111|\sout{1}111000|110110
\end{align*}

@@

\begin{align}
	D &=\begin{bmatrix}
		129\\ 
		14111238\\ 
		4685366
	    \end{bmatrix} 
        =
        \begin{bmatrix}
	  \ldots000000000000000010000001 \\
	  \ldots110101110101001000000110 \\
	  \ldots010001110111111000110110
	\end{bmatrix}
	\\\nonumber
	&=\begin{bmatrix}
\ldots\underbrace{000000}_{0}\underbrace{000000}_{0}\underbrace{000010}_{2}
\underbrace{000001}_{1} \\ 			 
\ldots\underbrace{110101}_{-11}\underbrace{110101}_{-11}\underbrace{001000}_{8}
\underbrace{000110}_{6} \\
\ldots\underbrace{010001}_{17}\underbrace{110111}_{23}\underbrace{111000}_{-8}
\underbrace{110110}_{-10}
	  \end{bmatrix}
\end{align}

\paragraph{Multiplication}

Let's now examine matrix multiplication. Consider the product $P = A \times
B^{t}$, where $B^t$ is the transpose of B. Thus the product becomes:

@@

\begin{equation}\label{eq:prodbs}
	P = AA \times BB =\begin{bmatrix}
			    4952\\ 
			    52841\\ 
			    14251
			  \end{bmatrix} 
        \times\times
	\begin{bmatrix}
	  36073 & 36303 & 686145
	\end{bmatrix}
\end{equation}

\begin{equation}
	P = AA \times BB = 
    	\begin{bmatrix}
\ldots\underbrace{0001}_{1}\underbrace{0011}_{3}\underbrace{0101}_{5}\underbrace
{1000}_{8}\\
\ldots\underbrace{1100}_{12}\underbrace{1110}_{14}\underbrace{0110}_{6}
\underbrace{1001}_{9}\\
\ldots\underbrace{0011}_{3}\underbrace{0111}_{7}\underbrace{1010}_{10}
\underbrace{1011}_{11}
	\end{bmatrix}
\end{equation}

\begin{small}
  \begin{equation}
	  \times\times \\
	  \begin{bmatrix}\mathsf{
\ldots\underbrace{00001}_{1}\underbrace{00011}_{3}\underbrace{00111}_{7}
\underbrace{01001}_{9}}  &\mathsf{
\ldots\underbrace{00001}_{1}\underbrace{00011}_{3}\underbrace{01110}_{14}
\underbrace{01111}_{15}} & \mathsf{
\ldots\underbrace{10100}_{20}\underbrace{11110}_{30}\underbrace{00010}_{2}
\underbrace{00001}_{1}}
	  \end{bmatrix}\nonumber
  \end{equation}
\end{small}

To calculate $P[1,1]$ which in uncompressed form is $1 \times 1 + 3 \times 3 + 5
\times 7 + 8 \times 9$, we first must extract the numbers from the bitstrings
and then proceed with the linear combination. The extraction will make use of
masks as described before. Let $tempA$ and $tempB$ store the values of blocks 1 
and 3 of AA[1,1] and BB[1,1]. Now we need to determine the size of the blocks
(bit-length) containing each element of the resulting matrix. As demonstrated
before, in the worst case, the product will have a bit-length which is the sum
of the bit-lengths of the operands. For this example this length is $4+5=9$,
however, we also have three additions for each element,adding a total of 3
extra bits, thus we end up with a bit-length of 12 for each element of the
product. Having determined the bit-length of the product, we can now \todo{we 
can - removi} do the actual calculations and store the results.

@@

Again, we can see that P is already in compressed form, which confirms that the
entire operation was conducted without decompressing the data.
